---
title: On the performance of for loops in R
author: Jose M Sallan
date: '2021-03-05'
slug: [for-loops-R]
categories:
  - R
tags:
  - R
  - optimization
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A common advice to R users is that we need to avoid `for` loops. This is because the basic data types of R are **vectors** (a variable in R is a vector of length one). Let's see a comparison of the performance of a vectorized operation versus a `for` loop.

## Filtering a vector

We need to find which positions of a numeric vector `vec` contain a number `num`. We may be tempted to do something like this:

```{r}
f1 <- function(vec, num){
   n <- length(vec)
  sol <- logical(n)
  k <- 1
  sol <- integer()
  for(i in 1:n){
    if(vec[i] == num){
      sol[k] <- i
      k <- k+1
    }
  }
  return(sol)
}
```

But as R is a vectorial language, we can obtain the same result doing:

```{r}
f2 <- function(vec, num) return(which(vec == num))
```

Let's build a large vector, test both functions on it and see if they bring the same result:

```{r}
set.seed(1313)
large_vector <- sample(0:9, 10000, replace = TRUE)
s1 <- f1(large_vector, 4)
s2 <- f2(large_vector, 4)
identical(s1, s2)
```

Let's examine the speed of both operations with the `rbenchmark` library:

```{r}
library(rbenchmark)
benchmark(f1(large_vector, 4), f2(large_vector, 4), columns=c('test', 'replications', 'elapsed', 'relative', 'user.self', 'sys.self'), replications = 100, order='elapsed')
```

We see that the function with a `for` loop is much slower than the vectorised function. This is empirical evidence advocating using vectorised functions when operating with vectors.

## Sum columns of a matrix

Let's now sum the columns of a very large matrix. We have three ways of doing that: with a `for` loop, with an `apply` loop or with a specific function.

### Three ways to do the job

To define a function that sums each column and puts the value into a vector component using a `for` loop:

```{r}
f3 <- function(m){
  n <- dim(m)[2]
  s <- numeric(n)
  for(i in 1:n) s[i] <- sum(m[,i])
  return(s)
}
```

A second possibility is to run a set of vectorized operations. We can do the same with an `apply` loop over columns:

```{r}
f4 <- function(m) return(apply(m, 2, sum))
```

Finally, we can use a built-in function called `colSums` that performs that task.

### Testing performance

Let's define a very large matrix `M`, and apply the three functions to it to obtain the sum of columns:

```{r}
M <- matrix(sample(1:100, 1000000, replace = TRUE), 1000, 1000)
s3 <- f3(M)
s4 <- f4(M)
s5 <- colSums(M)
```

The `s4` vector is integer, and the other two numeric. Let's check if they yield the same values:

```{r}
s4 <- as.numeric(s4)
identical(s3, s4)
identical(s4, s5)
identical(s3, s5)
```

Let's check the speed of each function:

```{r}
library(rbenchmark)
benchmark(f3(M), f4(M), colSums(M), columns=c('test', 'replications', 'elapsed', 'relative', 'user.self', 'sys.self'), order='elapsed')
```

The best performance is achieved by the built-in function. The `for` loop implementation seems to go faster than `apply` in this context.

## Compute the means of a list of vectors

Finally, we can consider the job of computing the means of a very large list of vectors. Let's define a very large list of vectors:

```{r}
vectors <- lapply(1:1000, function(x) sample(1:100, 10000, replace = TRUE))
```

We can compute the mean with a looping function over the list, using the vectorized `mean` function:

```{r}
looping_mean <- function(list){
  n <- length(list)
  means <- numeric(n)
  for(i in 1:n) means[i] <- mean(list[[i]])
  return(means)
}
```

An alternative for the `for` loop is the `sapply` function:

```{r, eval=FALSE}
sapply(vectors, mean)
```

Both functions return the same result:

```{r}
identical(looping_mean(vectors), sapply(vectors, mean))
```

Alternatively, we want to store the results on a list, instead of a vector:

```{r, eval=FALSE}
lapply(vectors, mean)
```

Let's examine the performance of each implementation:

```{r}
benchmark(looping_mean(vectors),
          sapply(vectors, mean),
          lapply(vectors, mean),
          columns=c('test', 'replications', 'elapsed', 'relative', 'user.self', 'sys.self'), 
          order='elapsed',
          replications = 100)
```

In this case, the three implementations have a similar performance. 

## When to avoid for loops in R

This small experiment show us when do we need to avoid for loops in R: when performing vectorized operations. As R is a vectorized language, **we must avoid looping across the components of a vector**, like when filtering a vector. This means that we must avoid looping in operations such as computing summarised vector statistics, or subsetting rows of a data frame. In other contexts, iterating functions of the `apply` family have a performance similar as `for` loops. The relative merits of each function seem to depend on each type of iteration.
